---
title: "Análises"
output: bookdown::html_notebook2
editor_options: 
  chunk_output_type: console
---

# Some theory 

## Analyzing and Interpreting Data From Likert-Type Scales

* Mean and standard deviation are hard do interpret in Linkert scales
* Use quantiles for Linkert scales
* Use contingency table and chi-square, Spearman rho assessment, U Mann-Whitney test
* Researchs recommend to use parametrics tests even for a Linkert scale
* When measuring subjective concepts, use Conbach alpha, kappa test and fatorial analysis

Where to read: [rcompanion](https://rcompanion.org/handbook/index.html)

**Parametric tests can be used Likert scale responses, but means are often limited value**

## Reporting and Interpreting Scores Derived from Likert-type Scales


# Used packages

```{r, message=FALSE}
library(readr)
library(readxl)
library(openxlsx)
library(stringr)
library(glue)
library(tidyverse)

# options to chunks
knitr::opts_chunk$set(echo = FALSE, out.width = "50%", fig.align = "center", cache=TRUE, warning = FALSE, message = FALSE)

# setting theme
theme_set(theme_minimal())

# Loading code
source("/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/scripts/functions/bar-graph.R")
source("/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/scripts/functions/contingency_table.R")
source("/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/scripts/functions/distribution-table.R")
source("/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/scripts/functions/grafico_2.R")
source("/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/scripts/functions/qui_quadrado_test.r")
source("/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/scripts/functions/resume-measures.R")
source("/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/scripts/functions/resume-measures-2.R")

# path directory
output <- stringr::str_interp('/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/output/')

# path to save figures
fig_path <- "/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/figures/"

# loading data
dados <- read_xlsx("/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/data/raw/Banco de dados atualizado15-03-211.xlsx")

# source("/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/scripts/tabela_grafico.R")

df <- dados |> 
  dplyr::mutate_at(dplyr::vars(dplyr::matches("^Q[4-6]$")), \(x) dplyr::recode(x,
                                                      `0` = 'Não',
                                                      `1` = 'Sim',
                                                      `2` = 'Não sei')) |> 
  dplyr::mutate_at(paste0("Q", 7:22), \(x) dplyr::recode(x,
                                       `0` = 'Sem preocupação',
                                       `1` = 'Pouca preocupação',
                                       `2` = 'Muita preocupação',
                                       `3` = 'Sem resposta')) |> 
  dplyr::mutate_at(paste0("Q", 23:26),
                   \(x) dplyr::recode(x,
                                      `0` = 'Não',
                                      `1` = 'Sim',
                                      `2` = 'Às vezes',
                                      `3` = 'Sem resposta')) |> 
  dplyr::mutate_at(paste0("Q", 27:39),
                   \(x) recode(x,
                               `0` = 'Nenhum dia',
                               `1` = 'Poucos dias',
                               `2` = 'Todos dias',
                               `3` = 'Todos dias e várias vezes')) |> 
  dplyr::mutate_at(paste0("Q", 40:43),
                   \(x) recode(x,
                               `0` = 'Não',
                               `1` = 'Sim',
                               `2` = 'Sem resposta',
                               `3` = 'Outros')) |>
  dplyr::mutate(raca = raça |>
           dplyr::recode('Não sei responder ou não quero responder' = 'Sem resposta',
                  'Parda' = 'Negra',
                  'Preta' = 'Negra'))

openxlsx::write.xlsx(df, file = "/home/gilberto/Dropbox/UFBA/consultoria/2021/consultoria-jualiana-adriana/data/processed/banco_dados.xlsx")
```

# Initial unidimensional analysis

## Cidade

### Bar graph

```{r grafico-barra-cidade}
grafico(dados, 'Cidade', 'Cidade', base = fig_path)
```

### Distribution table

```{r tabela-distribuicao-cidade}
tabela(dados, 'Cidade', 'Cidade', "Tabela de distribuição de frequências para variável Cidade.", base = output)
```

## Tipo de escola

### Bar graph

```{r grafico-barra-tipo-escola}
grafico(dados, 'Escola', "Tipo de Escola", base = fig_path)
```

### Distribution table

```{r tabela-distribuicao-tipo-escola}
tabela(dados, 'Escola', 'Tipo de Escola', "Tabela de distribuição de frequências para Tipo de Escola.", base = output)
```

## Gênero

### Distribuition tabela

```{r tabela-distribuicao-genero}
tabela(dados, 'Gênero', 'Gênero', 'Tabela de distribuição de frequências para Gênero', base = output)
```

### Bar graph

```{r grafico-barra-genero}
grafico(dados, 'Gênero', 'Gênero', base = fig_path)
```

## Idade

### Distribution table

```{r tabela-distribuicao-idade}
tabela(dados, 'Idade', 'Idade', 'Tabela de distribuição para Idade.', base = output)
```

### Bar graph

```{r grafico-barra-idade}
grafico(dados, 'Idade', 'Idade', base = fig_path)
```

## Raça

I have collapse `Negra`, `Parda` and `Preta` into `Negra`.

### Distribution table

```{r tabela-distribuicao-raca}
dados <- dados |> 
  mutate(raca = raça |>
           recode('Não sei responder ou não quero responder' = 'Sem resposta',
                  'Parda' = 'Negra',
                  'Preta' = 'Negra'))
tabela(dados, 'raca', 'Raça','Tabela de distribuição para Raça.', base = output)
```

### Bar graph

```{r grafico-barra-raca}
grafico(dados, 'raca', 'Raça', base = fig_path)
```

# Bidimensional analysis

In the dataset it refers to the following questions:

Q12) a Q22)

and we have used _linkert_ scale:

* 0 - Sem preocupação
* 1 - Pouca preocupação
* 2 - Muita preocupação
* 3 - Sem resposta

### Q12)

#### Distribution table

```{r tabela-distribuicao-q12}

tabela(df, 'Q12', 'Q12', 'Você está preocupado que demorasse muito para eu voltar à escola?', base = output)
```

### Bar graph

```{r grafico-barra-q12}
grafico(df, "Q12", "Q12 -- volta à escola", base = fig_path)
```

#### Resume measures unidimensional

```{r tabela-medidas-resumo-q12}
resume_measures(dados, "Q12", base = output)
```

#### Contingency table: Idade versus Questão 12)

```{r tabela-contingencia-q12}
contingency(df, "Idade", "Q12", "Idade", base = output)
```

#### Bar graph: Idade versus Questão 12)

```{r grafico-barra-idade-q12}
grafico_2(df, "Idade", "Q12", 'Idade', base = fig_path)
```

#### Chi-squared test: Idade versus Q12

```{r qui-quadrado-idade-q12}
df |> 
  qui_test("Idade", "Q12", base = output)
```


#### Resume Measures: Idade versus Questão 12

```{r medidas-resumo-idade-q12}
dados |> 
  resume_measures_2("Idade", "Q12", base = output)
```

#### Boxplot: Idade versus Q12

```{r}
ggplot(dados) +
  geom_boxplot(aes(x = as.factor(Idade), y = Q12)) +
  labs(x = "Idade", y = "Q12") +
  scale_y_continuous(breaks = -1:4, limits = c(-1,4))
```

#### Mann-Whitney test: Idade versus Q12

#### Dunn's Teste: Idade versus Q12

To do:

[] Boxplot
[] Man Whitney Tests
[] Organize Flux of results
